\documentclass[twoside,11pt]{article}
\usepackage[preprint]{jmlr2e}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows}
\usepackage{url}
\hypersetup{hidelinks}

\jmlrheading{0}{2026}{1-1}{7/26}{7/26}{TMLR-SUBMISSION}{Anonymous Authors}
\ShortHeadings{Benchmark-Specific Causal Structure in Session LLMs}{Anonymous Authors}
\firstpageno{1}

\title{Benign-Trained Session Language Models Recover Benchmark-Specific Causal Structure on Insider-Threat Benchmarks}

\author{\name Anonymous Authors \\
\addr Anonymous Institution}

\editor{Under Review by TMLR}

\begin{document}

\maketitle

\begin{abstract}% keep trailing percent for jmlr2e
We study whether a benign-trained session language model can be both
mechanistically analyzable and empirically meaningful on insider-threat-style
enterprise benchmarks. Our current paper-safe evidence centers on a
QLoRA-adapted Qwen3-8B model evaluated on two CERT benchmarks, r6.2 and r4.2.
After a July 2026 audit corrected detector evaluation and tightened the
mechanistic protocol, the detector story is mixed: fold-aligned remote detector
performance shows strong ROC and ranking behavior but weaker day PR-AUC than the
strongest local baselines. The mechanistic story is stronger. On r6.2, a
same-user-excluded token-level causal patching effect remains clearly positive
and the corresponding necessity effect is large. On r4.2, direct transfer of
the r6.2 mechanism fails, but a native r4.2 token mechanism can still be
rediscovered under the stricter same-user-excluded protocol. The resulting
scientific claim is not universal detector superiority or a universal sparse
insider-threat circuit. Instead, the current evidence supports a more specific
finding: benign-trained session LLMs can recover benchmark-specific causal
structure on insider-threat benchmarks, direct mechanism transfer can fail, and
native rediscovery can succeed.
\end{abstract}

\begin{keywords}
insider threat, anomaly detection, mechanistic interpretability, language
models, one-class learning, causal patching
\end{keywords}

\section{Introduction}
Enterprise anomaly detection is often evaluated only as a ranking problem, but
high-stakes insider-threat settings also require a usable answer to ``why did
the model fire?'' A detector that produces a score without any localized,
behaviorally meaningful mechanism is difficult to audit, hard to compare across
benchmarks, and easy to overclaim \citep{chandola2009anomaly}.

This draft focuses on a narrower question than universal anomaly detection:
whether a benign-trained session language model can recover stable,
benchmark-specific causal structure on insider-threat-style CERT benchmarks. The
current remote branch is a QLoRA-adapted Qwen3-8B session model trained only on
benign users. Its evaluation has two parts. First, a detector benchmark asks
how the remote score behaves under the same leave-one-malicious-user-out folds
used by the local baselines. Second, a mechanistic benchmark asks whether
sparse token feature sets show causal sufficiency and necessity under active
controls and after same-user exclusions. The model family itself is still a
transformer-based sequence model \citep{vaswani2017attention}.

The current evidence supports three points. First, benign-only remote training
is valid one-class training under the implemented protocol. Second, the remote
mechanistic branch is genuinely positive on r6.2 after the stricter same-user
repair. Third, r4.2 rejects a universal transfer story but still supports
native rediscovery of a smaller remote mechanism. What the evidence does not
support is a strong detector-superiority claim on CERT.

\input{figures/pipeline.tex}

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=1.3em]
\item We provide a repaired, paper-safe CERT evaluation policy that separates
fold-aligned detector benchmarking from same-user-excluded mechanistic
evaluation.
\item We show that the remote token mechanism on r6.2 remains positive after the
same-user donor/match audit.
\item We show that direct token-mechanism transfer from r6.2 to r4.2 fails, but
a native r4.2 mechanism can still be recovered.
\item We show that detector competitiveness and mechanistic interpretability
need not move together: the mechanistic claim is stronger than the detector
claim in the current repaired evidence.
\end{enumerate}

\section{Setup and Evaluation Protocol}
\subsection{Remote model and data protocol}
The remote branch is a benign-trained Qwen3-8B QLoRA session model. Positive
users are assigned to the remote evaluation split, while training and validation
use benign users only. This makes the remote branch a one-class detector in the
sense that the model is never trained on malicious examples.

\subsection{Datasets}
We focus on two CERT benchmarks. The r6.2 branch has 70 remote positive
receiver-days in the current mechanistic evaluation. The r4.2 branch is larger,
with 1309 remote positive receiver-days in the repaired native-mechanism
evaluation. Both datasets also have local Magnolia baselines and local
session-AE mechanistic comparators \citep{certdataset}.

\subsection{Detector benchmark}
After the July 2026 audit, remote detector rows are considered valid only when
they use full-population scoring on \texttt{all.jsonl} followed by fold-aligned
evaluation on the same leave-one-malicious-user-out test-user construction as
the local baselines. This protocol fixes the earlier split-filter bug that had
made the old \texttt{detector\_metrics/} artifacts too optimistic.

\subsection{Mechanistic estimands}
The causal branch measures a token-level repair advantage:
top sparse feature sets are patched into positive receivers and compared against
an active control set. The necessity branch measures whether ablating the top
feature set hurts anomalous examples more than ablating the active control.
Final mechanistic rows are taken only from the repaired same-user-excluded
runs.

\input{figures/story.tex}

\section{Results}
\subsection{Detector results are mixed under the corrected protocol}
The corrected detector table is fold-aligned and therefore scientifically usable
for remote-vs-local comparison. It changes the story materially: the remote
detector has strong ROC and some useful user-ranking behavior, but it is not
stronger than the best local Magnolia baselines on day PR-AUC, including the
Deep SVDD reference baseline \citep{ruff2018deepsvdd}.

\input{tables/cert_detector_comparison.tex}

The consequence is straightforward. Detector evidence should stay in the paper,
but not as the headline claim. The strongest safe reading is that the remote
detector remains nontrivial and behaviorally structured, not that it dominates
the local baselines.

\subsection{r6.2 retains a strong remote mechanism after repair}
The repaired r6.2 same-user-excluded causal run remains clearly positive for the
main organization-context rows, and the matched necessity effect remains large.
The repaired remote-vs-local day-level comparison also shows that the remote
causal effect remains larger than the local session-AE comparator scale.

\subsection{r4.2 rejects direct transfer but supports native rediscovery}
Transferred layer-18 r6.2 token configurations remained negative on r4.2, so
the data do not support a universal transferable sparse circuit. However, the
native r4.2 search found a positive layer-26 token branch, and that branch
remained positive under the stricter same-user-excluded recovery. The native
necessity result is weaker than r6.2 and is strongest for \texttt{dept\_role}
and \texttt{role}.

\input{tables/cert_mechanistic_summary.tex}

\subsection{The paper claim is now narrower but cleaner}
After the recovery work, the project has a more defensible paper shape:
mechanistic evidence is the center, detector performance is a qualified
background result, and direct transfer failure becomes part of the scientific
finding rather than a flaw to hide.

\input{tables/claim_status.tex}

\section{Discussion}
The repaired evidence supports a paper about benchmark-specific causal structure
in benign-trained session language models. That is a meaningful result on its
own. It says that the learned signal is not merely a detector score but can be
localized to sparse token-level mechanisms, even if those mechanisms are not
portable across benchmarks without rediscovery.

This also clarifies what should \emph{not} be claimed. The current detector
results do not justify a strong ``better detector'' paper. The current
mechanistic results also do not justify a universal insider-threat circuit.
Instead, the strongest supported message is that the same modeling family can
recover different, benchmark-specific mechanisms under a common remote
pipeline.

\section{Limitations}
This is still a first draft and not a final submission package. Two limitations
matter immediately. First, the detector story is weaker than the mechanistic
story and should be framed that way. Second, this draft still uses generated
tables and simple placeholder figures rather than the polished final visual
package that a submission would need.

\section{Broader Impact Statement}
Insider-threat detection can reduce harm from credential misuse, data exfiltration,
and risky internal behavior, but it also creates surveillance risks. A model
that is operationally deployed without strong governance could be used to
escalate normal employee variation, over-monitor specific workers, or justify
opaque automated action. The current work modestly improves that situation by
favoring benign-only training, explicit evaluation policies, and mechanistic
auditing over purely opaque score maximization. Even so, none of these results
should be interpreted as a justification for autonomous disciplinary action.
Human review, documented appeal processes, and careful organizational policy are
required for any responsible deployment.

\acks{This anonymized draft omits acknowledgements for submission. The final
camera-ready version should add funding, disclosure, and acknowledgements as
required by the venue.}

\bibliography{references}

\end{document}
